---
title: "ADA Final Project"
author: "Arifa Nayab"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
##installing and loading required packeges 
```{r}
library(haven)
library(dplyr)
library (survey)
library(ggplot2)
library(gtsummary)
pacman::p_load(odds.n.ends, blorr, lmtest, car, broom, tidyverse, readr, jtools) 
```

##importing BRFSS data from 2022 and 2023
```{r}
brfss2022 <- read_xpt("C:/Users/nayab/OneDrive/Desktop/ADA.final.project/LLCP2022.XPT") 
```

```{r}
brfss2023 <- read_xpt("C:/Users/nayab/OneDrive/Desktop/ADA.final.project/LLCP2023.XPT")
```
##adding year variable to the datasets 
```{r}
brfss22 <- brfss2022 %>% mutate(year = 2022)
brfss23 <- brfss2023 %>% mutate(year = 2023) #to account for survay year and to look for changes in trends in each year
```

##Renaming Race and Insurance variable so that they are labbled with the same name in both datasets for easy merging 
```{r}
brfss23 <- brfss23 %>% rename(`_RACE1` = `_RACE`)
brfss23 <- brfss23 %>% rename(`PRIMINSR` = `PRIMINS1`)
```

##combining 2022 and 2023 datasets into one
```{r}
brfss_all <- bind_rows(brfss22, brfss23) #combining BRFSS data from 2022 and 2023 into one file
```

## create new dataset with only variables of interest 
```{r}
brfss_selected <- brfss_all %>%
  select("_LLCPWT", "_PSU", "_STSTR", "_AGE_G", "ECIGNOW2", "CHCCOPD3", "LADULT1", "SEXVAR","_RACE1", "_EDUCAG", "INCOME3", "PRIMINSR", "_BMI5CAT", "ASTHMA3", "_SMOKER3")

head(brfss_selected) #to check if all the required variables are in the new dataset
```

```{r}
names(brfss_selected)
```
  
##rename variables for easy understanding 
```{r}
brfss_selected <- brfss_selected %>%
  rename("Ecig use"= ECIGNOW2, "Respiratory complication" = CHCCOPD3, "Age 18+" = LADULT1, "Sex" = SEXVAR, "Race" = '_RACE1', "Education Level" = '_EDUCAG', "Income" = INCOME3, "Insurance" = PRIMINSR, "BMI_cat" = '_BMI5CAT', "Age" = `_AGE_G`)
```

```{r}
sapply(brfss_selected, typeof) #to check the type of variable and if renaming worked 
```

#recoding variables for analysis 
#variable 1 respiratory complications 
```{r}
table(brfss_selected$`Respiratory complication`, useNA = "always")

# recode as binary outcome variable and factor it
brfss_selected <- brfss_selected %>%
  mutate(Complications_binray = case_when(`Respiratory complication` == "1" ~ 1,  #yes
                                          `Respiratory complication` == "2" ~ 0), #NO
         Complications_binray = factor(Complications_binray, levels=c (0,1), labels=c("No", "Yes"))) # make it a factor variable

# check to make sure re-categorization worked as expected
table(brfss_selected$`Respiratory complication`, brfss_selected$Complications_binray, useNA="always")
```
#variable 2 E-cigarette use
```{r}
table(brfss_selected$`Ecig use`, useNA = "always")

# recode as categorical variable and factor it
brfss_selected <- brfss_selected %>%
  mutate(Eciguse_cat = case_when(`Ecig use` == "1" ~ 0,             #Never used
                                 `Ecig use` %in% c("2","3") ~ 1,    #Current user
                                 `Ecig use`== "4" ~ 2),             #former user
         Eciguse_cat = factor(Eciguse_cat, levels=c(0,1,2), labels=c("Never used", "Curent user", "former user"))) # make it a factor variable

# check to make sure re-categorization worked as expected
table(brfss_selected$`Ecig use`, brfss_selected$Eciguse_cat, useNA="always")
```
```{r}
table(brfss_selected$ASTHMA3, useNA = "always")

# recode as binary variable and factor it
brfss_selected <- brfss_selected %>%
  mutate(Asthma_binary = case_when(ASTHMA3 == 1 ~ 1,  #Yes
                                   ASTHMA3 == 2 ~ 0), #No
         Asthma_binary = factor(Asthma_binary, levels=c(0,1), labels=c("No", "Yes"))) # make it a factor variable

# check to make sure re-categorization worked as expected
table(brfss_selected$ASTHMA3, brfss_selected$Asthma_binary, useNA="always")
```

#income level

```{r}
table(brfss_selected$`Income`, useNA = "always")

# recode as categorical variable and factor it
brfss_selected <- brfss_selected %>%
  mutate(Income_cat = case_when(`Income` %in% c("1","2", "3", "4") ~ 0,   #Low income
                                `Income` %in% c("5","6", "7") ~ 1,        #Middle income
                                `Income` %in% c("8","9","10","11")  ~ 2), #High income
         Income_cat = factor(Income_cat, levels=c(0,1,2), labels=c("Low income", "Middle income", "High income"))) # make it a factor variable

# check to make sure re-categorization worked as expected
table(brfss_selected$`Income`, brfss_selected$Income_cat, useNA="always")
```
##Insurance 

```{r}
table(brfss_selected$`Insurance`, useNA = "always")

# recode as categorical variable and factor it
brfss_selected <- brfss_selected %>%
  mutate(Insurance_cat = case_when(`Insurance` %in% c("1","2") ~ 0,   #Private
     `Insurance` %in% c("3","4","5","6","7","8","9","10") ~ 1,        #Government provided
                                `Insurance` %in% c("11")  ~ 2), #No insurance
         Insurance_cat = factor(Insurance_cat, levels=c(0,1,2), labels=c("Private insurance", "Government provided insurance", "No insurance"))) # make it a factor variable

# check to make sure re-categorization worked as expected
table(brfss_selected$`Insurance`, brfss_selected$Insurance_cat, useNA="always")
```
##Sex

```{r}
brfss_selected <- brfss_selected %>%
  mutate(Sex=case_when(Sex == 1 ~"Male",
                       Sex == 2 ~"Female"))

table(brfss_selected$Sex, useNA = "always")
```
##Age
```{r}
brfss_selected <- brfss_selected %>% 
  filter(`Age 18+` == 1) %>% 
  mutate(`Age 18+` = case_when(`Age 18+` == 1 ~ "yes",
                               `Age 18+` == 2 ~ "no"))
table(brfss_selected$`Age 18+`, useNA = "always")
```

```{r}
names(brfss_selected)
```

```{r}
#Cross-checking to see if the re-coding worked or not 
sapply(brfss_selected[, c("Eciguse_cat", "Complications_binray", "Age 18+", "Sex", "Race", "Education Level", "Income_cat", "Insurance_cat", "BMI_cat","Asthma_binary","_SMOKER3", "Age")],
  table,
  useNA = "ifany")
```


##making Didn't respond, Don't know/Not sure as NAs 
```{r}
brfss_selected <- brfss_selected %>%
  mutate(
    across(
      c(`_SMOKER3`, `Education Level`),
      ~ replace(., . %in% c(9, 77, 99), NA)
    )
  )
```

###Dropping row with missing data for analysis variables (for complete case analysis)
```{r}
analysis_vars <- c(
  "Eciguse_cat", "Complications_binray", "Age 18+", "Sex", "Race", "Education Level", "Income_cat", "Insurance_cat", "BMI_cat","Asthma_binary","_SMOKER3","Age")

brfss_selected <- brfss_selected %>%
  drop_na(all_of(analysis_vars))   

#testing to check the variables

table(brfss_selected$Eciguse_cat, useNA = "ifany")
table(brfss_selected$Complications_binray, useNA = "ifany")
table(brfss_selected$`_SMOKER3`, useNA = "ifany")
table(brfss_selected$`Age 18+`, useNA = "ifany")
table(brfss_selected$Sex, useNA = "ifany")
table(brfss_selected$Race, useNA = "ifany")
table(brfss_selected$`Education Level`, useNA = "ifany")
table(brfss_selected$Income_cat, useNA = "ifany")
table(brfss_selected$Insurance_cat, useNA = "ifany")
table(brfss_selected$BMI_cat, useNA = "ifany")
table(brfss_selected$Age, useNA = "ifany")
table(brfss_selected$Asthma_binary, useNA = "ifany")
```


##re-coding to label catagories  

```{r}

#for smoking status 
brfss_selected <- brfss_selected %>%
  mutate(
    `_SMOKER3` = factor(`_SMOKER3`,   
                  levels = c(1,2,3,4),
                  labels = c("Current everyday smoker","current somedays smoker","Former Smoker","Never smoked")))

#for race variable 
brfss_selected <- brfss_selected %>%
  mutate(
    Race = factor(Race,   
                  levels = c(1,2,3,4,5,7,8),
                  labels = c("Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic American Indian or Alaskan Native", "Non-Hispanic Asian","Non-Hispanic Native Hawaiian or other Pacific Islander","Non-Hispanic Multiracial","Hispanic")))

#for education level 
brfss_selected <- brfss_selected %>%
  mutate(
    `Education Level` = factor(`Education Level`,   
                  levels = c(1,2,3,4),
                  labels = c("Did not graduate High School","Graduated High School", "Attended College or Technical School", "Graduated from College or Technical School")))

#for BMI
brfss_selected <- brfss_selected %>%
  mutate(
    BMI_cat = factor(BMI_cat,   
                  levels = c(1,2,3,4),
                  labels = c("Underweight","Normal Weight","Overweight","Obese")))
#for age catagory
brfss_selected <- brfss_selected %>%
  mutate(
    Age = factor(Age,
                     levels = c(1,2,3,4,5,6),
                     labels = c("18–24","25–34","35–44","45–54","55–64","65+")))
```



```{r}
names(brfss_selected)
```

## Make a staked bar plot to visualize whether there is a difference in the respiratory complications based on E-Cig use.

```{r}
  ggplot(brfss_selected, aes(x = Eciguse_cat, fill = Complications_binray)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "E-cigarette Use",
    y = "Percent",
    fill = "Respiratory Complications"
  ) +
  theme_bw()
```
```{r}
table(brfss_selected$Complications_binray)

```




##Binary logistic Regression Analysis (univariate analysis)
```{r}
model10<- glm(Complications_binray ~ Eciguse_cat, data=brfss_selected, family="binomial")
summary(model10) # get log results

myoutput <- odds.n.ends(model10 ) # ORs and CIs using odds.n.ends function
myoutput
```
The odds of having respiratory complications among those who are current e-cigarette users is 2.89 times higher compared to those who Never used e-cigarettes

The odds of having respiratory complications among those who are former e-cigarette users is 2.07 time higher compared to those who never used e-cigarettes

```{r}
# adjust for age and run model and get results
model11 <- glm(Complications_binray ~  Eciguse_cat + Age, data=brfss_selected, family="binomial")

odds.n.ends(model11)

# compare coefficients between models 10 and 11
compareCoefs(model10, model11)
```
##Interpretation 
The odds of having respiratory complications among those who are current e-cigarette users is 3.78 (95% CI 3.39-4.22) times higher compared to those who Never used e-cigarettes when adjusted for age

The odds of having respiratory complications among those who are former e-cigarette users is 2.20 (95% CI 2.10-2.30) time higher compared to those who never used e-cigarettes when adjusted for age


```{r}
model_full <- glm(
  Complications_binray ~ Eciguse_cat +
   Age  + Sex + Race + `Education Level` +
    Income_cat + Insurance_cat + BMI_cat +
    Asthma_binary + `_SMOKER3`,
  data = brfss_selected,
  family = binomial()
)

odds.n.ends(model_full)

# compare coefficients between models 11 and model_full
compareCoefs(model11, model_full)

```


The odds of having respiratory complications among those who are current e-cigarette users is 1.70 (95% CI 1.50-1.92) times higher compared to those who Never used e-cigarettes, after adjusted for age, sex, race, education level, income level, insurance status, MBi, Asthma status and smoking status. 

The odds of having respiratory complications among those who are former e-cigarette users is 1.34 (95% CI 1.27-1.41) time higher compared to those who never used e-cigarettes, after adjusted for age, sex, race, education level, income level, insurance status, MBi, Asthma status and smoking status.


## Look at assumptions of multicollinearity using the vif function from the car package

```{r}
vif(model_full)

```
```{r}
# Plot Cooks's distance
plot(model_full, which = 4, id.n = 3, col="red") # which = 4 gives Cook's distance calculations
abline(h=1, col="red", lty=2)  # Add a threshold line
```

##Testing for assumptions (Binary logistic Regression)
1. Binary dependent variable: Respiratory conditions is a binary outcome with Yes/No possibilities 
2. Independence of observation: Independence of observation is violated in BRFSS data and survey weights are usually applied to make sure the assumption is not violated (survey weights are not applied in this analysis). 
3. Linearity of logit - There are no continuous independent variables. 
4. No multicollinearity - checked using VIF and the results show that there is no multicollianrity 
5. Sufficient sample size - BRFSS sample size is sufficient 
6. No extreme outliers - No extreme outliers based on cooks distance plot

##Secondary objective:Determine whether smoking status modifies the relationship between e-cigarette use and respiratory complications.
```{r}
#Model 1 for main effects only
model_main <- glm(
  Complications_binray ~ Eciguse_cat +
    Age + Sex + Race + `Education Level` +
    Income_cat + Insurance_cat + BMI_cat +
    Asthma_binary + `_SMOKER3`,
  data = brfss_selected,
  family = binomial()
)

#Model 2 for added ace_exposure * sex interaction
model_int <- glm(
  Complications_binray ~ Eciguse_cat*`_SMOKER3`+
    Age + Sex + Race + `Education Level` +
    Income_cat + Insurance_cat + BMI_cat +
    Asthma_binary,
  data = brfss_selected,
  family = binomial()
)

## Likelihood ratio test for interaction (effect modification)
lrtest(model_main, model_int)
```
#Interpretation (hypothesis test)

H₀: There is no interaction between e-cigarette use and smoking status 

H₁: There is an interaction between e-cigarette use and smoking status.

p < 0.05 from likelihood ratio test, we reject H₀.

Hence, There is evidence that smoking status modifies the association between e-cigarette use and respiratoey complications.  
(LR χ²(6) = 14.6, p = 0.023). 

```{r}
tab1_vars <- c(
  "Complications_binray", 
  "Age",  
  "Sex", 
  "Race",
  "Education Level",
  "Income_cat",
  "Insurance_cat", 
  "BMI_cat",
  "Asthma_binary",
  "_SMOKER3"
)

table1_exposure <- brfss_selected %>%
  select(Eciguse_cat, all_of(tab1_vars)) %>%
  tbl_summary(
    by = Eciguse_cat,
    statistic = all_categorical() ~ "{n} ({p}%)",
    missing = "no"
  ) %>%
 
  bold_labels()

table1_exposure

```

